\documentclass[12pt,PhD]{muthesis}
\bibliographystyle{styles/vancouver}
\usepackage{cite}
\usepackage{graphics}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{url} % typeset URL's reasonably
\usepackage{listings}

\newcommand{\unit}[2]{\ensuremath{#1\,\text{#2}}}
\newcommand{\cms}[1]{\ensuremath{#1\,\text{cms}^{\text{-1}}}}

\newcommand{\ms}[1]{\unit{#1}{ms}}
\newcommand{\msrange}[2]{\ensuremath{#1\text{--}#2\,\text{ms}}}
\newcommand{\mm}[1]{\unit{#1}{mm}}
\newcommand{\mv}[1]{\unit{#1}{mV}}
\newcommand{\degr}[1]{\ensuremath{#1^{\circ}}}

\newcommand{\supers}[1]{\ensuremath{^{#1}}}
\newcommand{\subs}[1]{\ensuremath{_{#1}}}
\newcommand{\ii}[1]{\ensuremath{\text{\emph{I}}_{\text{#1}}}}
\newcommand{\iip}[1]{\ensuremath{\text{\emph{I}}'_{\text{#1}}}}
\newcommand{\g}[1]{\ensuremath{\text{\emph{g}}_{\text{#1}}}}

\newcommand{\apd}[1][90]{APD\subs{#1}}
\newcommand{\apdr}[1][90]{APD\emph{r}{\scriptsize,#1}}


\newcommand{\nothree}{\ensuremath{\text{NO}_{3^{\text{-}}}}}

% picture predefines
\setlength{\unitlength}{1mm}

\newsavebox{\resistor}
\savebox{\resistor}(10,2)[bl]{
\put(0, 1){\line(1, 0){1}}
\put(1, 0){\line(0, 1){2}}
\put(1, 0){\line(1, 0){8}}
\put(1, 2){\line(1, 0){8}}
\put(9, 0){\line(0, 1){2}}
\put(9, 1){\line(1, 0){1}}
}

\newsavebox{\vresistor}
\savebox{\vresistor}(2, 10)[bl]{
\put(1, 0){\line(0, 1){1}}
\put(0, 1){\line(1, 0){2}}
\put(0, 1){\line(0, 1){8}}
\put(2, 1){\line(0, 1){8}}
\put(0, 9){\line(1, 0){2}}
\put(1, 9){\line(0, 1){1}}
}

\newsavebox{\cell}
\savebox{\cell}(10,10)[bl]{
\put(0, 0){\line(0, 1){10}}
\put(10, 0){\line(0, 1){10}}
\put(0, 0){\line(1, 0){10}}
\put(0, 10){\line(1, 0){10}}
}

% algorithmic package
\renewcommand{\algorithmiccomment}[1]{// #1}

\begin{document}

    \title{Developing Realistic Models of the Atrium and the P-Wave ECG}
    \author{Jonathan D. Stott}
    \principaladviser{Henggui Zhang}
    \school{School of Physics and Astronomy}

    \beforeabstract
    \prefacesection{Abstract}
    \singlespace
    \input{chapters/abstract}
    \onehalfspace
    \afterabstract
    \prefacesection{Acknowledgements}
I would like to thank my supervisor, Prof. Henggui Zhang of the Biological Physics
group at Manchester, for his support and guidance throughout my thesis.
Without his help, this work would not have been completed.

I would also like to thank all of the members of the Biological Physics group,
past and present, for their conversations on the thesis and on the topics around
it (and everything else).
Through them I have learned of papers I might have missed and considered avenues
of investigation that otherwise I might not have done.
This gratitude is extended to the members of the Cardiovascular Physiology
group at Manchester, especially Prof. Mark Boyett who leads the group and who
has been a source of valuable discussion on topics all around the field of
cardiac electrophysiology.

Another debt of gratitude is owed to Doctors Clifford Garret, Derek Rowlands and
Gwilym Morris.
Their discussions about the ECG in general and the ones produced in this thesis
in specific were helpful as I refined the model.

My parents, David and Judith Stott and my sister, Joanna.
I'm not sure how to thank them for all their love and support as I have worked
on this thesis.

My friends, both local and remote.
For occasional discussions about programming, my work and anything else that was
on my mind.
They've helped me, probably in more ways than they know.

I would like to thank EPSRC for providing the funding which enabled me to
complete my studies.

Finally, I would like to thank all those who saw drafts of this thesis.
Without them, the final text would have more errors.
Though, of course, any errors  which remain are my own.

\prefacesection{The Author}

Jonathan Stott graduated with an Master of Physics (2:1) degree from the
University Of Manchester in 2005.
This thesis represents his first major research project.
A smaller research project in the field was completed as part of his Master's
course, which was presented at the Physiological Society meeting in Bristol
(2005).


    \afterpreface
    

   \include{chapters/introduction}
   \include{chapters/toolkit}
   \include{chapters/atrium_modelling}
   \include{chapters/body_surface_potential}
   \include{chapters/the_forward_problem}
   \include{chapters/conclusions}

%    \bibliography{refs}
    \bibliography{refs_full}

\end{document}
