\chapter{Physiological and Mathematical Background}

The heart is one of the most important organs in the human body.
It is responsible for pumping blood around the arteries and veins of the human
body to all of the organs within.
The blood carries all of the oxygen, energy and other substances needed for the
body to continue to live to all parts of the body.
It also bears away all of the carbon dioxide and other waste products so that
they can be removed.

The heart does this via rhythmic and synchronised contractions.
The contractions are, in the healthy heart, initiated from one place and
synchronised via electrical signals conducted through the heart.
This conduction is aided by a number of specialised cell types.
The conduction, and the cellular electrophysiology which underpins it, can be
modelled mathematically and the equations solved on computers.

This chapter provides the physiological background necessary to understand the
functioning of the heart.
It explains the theory behind and the normal form of the electrocardiograph.
It also introduces the concepts behind mathematical models of cardiac cells, the
heart and the ECG.
Finally, it discusses existing modelling toolkits, why such toolkits exist and
the limitations of toolkits in general and in specific.

\input{chapters/background/the_heart}
\input{chapters/background/activation}
\input{chapters/background/the_electrocardiogram}
\input{chapters/background/mathematical_models}
\input{chapters/background/the_forward_problem}
\input{chapters/background/toolkit}
